GMSE: an R package for generalised management strategy evaluation

### 2017-11-08 notebook/ms/SI1.Rmd
More on fitness function for managers

Manager fitness function
--------------------------------------------------------------------------------

User fitness function
--------------------------------------------------------------------------------

>>>>

Fitness function for managers
--------------------------------------------------------------------------------

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above). Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}$ is positive. To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation for the individual $i$,

$$ D_{i} = (\Delta A_{i} - U^{m}_{res})^2. $$

The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that does not get closer to the `manage_target` abundance. Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting this value from $D_{i}$ for each individual,

$$ F_{i} = D_{max} - D_{i}. $$

We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used). We now explain in more detail how individuals in the genetic algorithm calculate their actions will affect $\Delta A_{i}$. Below explains calculations for a single individual $i$ in the genetic algorithm, so we will drop the use of subscripts $i$ to refer to individuals; instead we use subscripts $j$ to refer to actions (e.g., scaring, culling, etc.).

To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C_{old}^{j}$). This information is collected from `ACTION` and `COST` arrays. The individual then needs to predict how their policy (i.e., the costs that they set users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$). To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy. The predicted total number of a particular action $j$ performed is thereby calculated as,

$$ X_{j}^{new} =  (X_{j}^{old} + 1) \frac{C^{old}_{j}}{C^{new}_{j}}. $$

The variable $C_{new}^{j}$ defines the new cost set by the individual for action $j$. A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step. Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no affect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.

The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `scaring = FALSE`).

Fitness function for users
--------------------------------------------------------------------------------
